\begin{abstract}
This report presents the comprehensive design, implementation, and performance analysis of advanced control strategies for double-inverted pendulum (DIP) stabilization---a canonical benchmark for underactuated nonlinear systems. Six controller variants are developed and compared: classical SMC with boundary layer, super-twisting algorithm (STA-SMC) for chattering reduction, adaptive SMC with online gain tuning, hybrid adaptive STA-SMC combining both techniques, model predictive control (MPC) for constraint-aware optimization, and swing-up SMC for large-angle control. Controller gains are systematically optimized using particle swarm optimization (PSO) with a multi-objective cost function balancing settling time, overshoot, energy consumption, and chattering amplitude.

Comprehensive Monte Carlo benchmarks ($N=100$ trials per controller) demonstrate: (1) MPC achieves fastest settling time (1.48~s) with 1.2\% overshoot and zero chattering through predictive optimization, (2) PSO-optimized hybrid adaptive STA-SMC provides 40\% faster settling than classical SMC with 70\% chattering reduction and superior robustness (12\% degradation under $\pm$30\% parameter uncertainty), (3) Swing-up controller achieves 94\% success rate for large-angle transitions from 180$^\circ$ hanging position. Statistical validation confirms all performance improvements with 95\% confidence intervals.

Beyond controller design, this report documents the complete production-grade software architecture: 352 source files across 16 subsystems, 237 test files (89\% coverage), comprehensive optimization infrastructure (5 algorithms), hardware-in-the-loop validation, and safety/monitoring systems. Exploratory research investigations (7 phases) into hybrid controller stability are documented, providing transparency into the iterative development process. All implementations are validated through high-fidelity simulation using fourth-order Runge-Kutta integration and provided as open-source software (91 scholarly citations) for reproducibility and extension.

\vspace{0.5em}
\noindent\textbf{Keywords}: Sliding mode control, model predictive control, double-inverted pendulum, particle swarm optimization, super-twisting algorithm, swing-up control, software architecture, robust control
\end{abstract}
