%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

This textbook would not have been possible without the support and contributions of many individuals and organizations.

First, I thank my academic advisors [Names] for their guidance, patience, and encouragement throughout this research journey. Their insights shaped both the theoretical foundations and practical implementations presented in this work.

I am grateful to [Institution/Department] for providing the computational resources and research environment that enabled this work. The [Grant/Fellowship Name], awarded by [Funding Agency], provided essential financial support.

Special thanks to the reviewers and early readers who provided invaluable feedback on draft chapters: [Names]. Their detailed comments significantly improved the clarity and rigor of the presentation.

I acknowledge the open-source community, particularly the developers of NumPy, SciPy, Matplotlib, PySwarms, and Numba. These libraries form the foundation of the Python implementations described in this textbook. The pandas, pytest, and Streamlit teams also deserve recognition for enabling data analysis, testing, and interactive visualization.

Thanks to my colleagues and fellow researchers who engaged in discussions, shared insights, and contributed to the collaborative environment that fostered this work: [Names].

Finally, I thank my family [Names] for their unwavering support, understanding, and patience during the many hours dedicated to writing, coding, and revising this textbook.

Any errors or omissions in this work are solely my responsibility.

\vspace{1cm}

\noindent [Author Name] \\
January 2026
